# Labels {#labels}

## Introduction

In this chapter, we will learn to:

- add title and subtitle to the plot
- modify axis labels
- modify axis range
- remove axis
- format axis

```{r axis1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Basic Plot

Let us start with a simple scatter plot. We will continue to use the mtcars 
data set and examine the relationship between displacement and miles per gallon
using `geom_point()`.

```{r axis2, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg))
```

## Title & Subtitle

There are two ways to add title to a plot:

- `ggtitle()`
- `labs()`

### ggtitle()

Let us explore the `ggtitle()` function first. It takes two arguments:

- label: title of the plot
- subtitle: subtitle of the plot

```{r axis3, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  ggtitle(label = 'Displacement vs Mileage', subtitle = 'disp vs mpg')
```

## Axis Labels

You can add labels to the axis using:

- `xlab()`
- `ylab()`
- `labs()`

```{r axis4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  xlab('Displacement') + ylab('Miles Per Gallon')
```

## Labs

The `labs()` function can be used to add the following to a plot:

- title
- subtitle
- X axis label
- Y axis label

```{r axis5, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  labs(title = 'Displacement vs Mileage', subtitle = 'disp vs mpg', 
    x = 'Displacement', y = 'Miles Per Gallon')
```

## Axis Range

In certain scenarios, you may want to modify the range of the axis. In ggplot2, 
we can achieve this using:

- `xlim()`
- `ylim()`
- `expand_limits()`


- `xlim()` and `ylim()` take a numeric vector of length 2 as input
- `expand_limits()` takes two numeric vectors (each of length 2), one for each axis 
- in all of the above functions, the first element represents the lower limit and the second element represents the upper limit

### X Axis

In the below example, we limit the range of the X axis between `0` and `600` using `xlim`.

```{r axis6, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  xlim(c(0, 600))
```

### Y Axis

Let us limit the range of the Y axis between `0` and `40`.

```{r axis7, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  ylim(c(0, 40))
```

### Expand Limits

Let us use `expand_limits()` to limit the range of both the X and Y axis. The 
first input is the range for the X axis and the second input for the Y axis. In
both the cases, we use a numeric vector of length 2 to specify the lower and 
upper limit.

```{r axis8, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  expand_limits(x = c(0, 600), y = c(0, 40))
```

### Remove Axis Labels

If you want to remove the axis labels all together, use the `theme()` function. 
It allows us to modify every aspect of the theme of the plot. Within `theme()`, 
set the following to `element_blank()`.

- `axis.title.x`
- `axis.title.y`

`element_blank()` will remove the title of the X and Y axis.

```{r axis9, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank())
```

## Format Title & Axis Labels

To format the title or the axis labels, we have to modify the theme of the plot 
using the `theme()` function. We can modify: 

- color
- font family
- font face
- font size 
- horizontal alignment 
- and angle

In addition to `theme()`, we will also use `element_text()`. It should be used
whenever you want to modify the appearance of any text element of your plot.

### Color

In the below example, we use the `color` argument within `element_text()` to
modify the color of the title of the plot to `'blue'`.

```{r axis10, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + ggtitle('Diaplacement vs Mileage') +
  theme(plot.title = element_text(color = 'blue'))
```

### Font Family

Let us change the font family of the plot title to `'Arial'` by using the
`family` argument.

```{r axis11, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + ggtitle('Diaplacement vs Mileage') +
  theme(plot.title = element_text(family = 'Arial'))
```

### Font Face

The font face can be any of the following:

- `plain` 
- `bold`
- `italic` 
- `bold.italic`

The `face` argument can be used to modify the font face of the title of the 
plot. 

```{r axis12, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + ggtitle('Diaplacement vs Mileage') +
  theme(plot.title = element_text(face = 'bold'))
```

### Font Size

The size of the title of the plot can be modified using the `size` argument.

```{r axis13, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + ggtitle('Diaplacement vs Mileage') +
  theme(plot.title = element_text(size = 8))
```

## Horizontal Alignment

To modify the horizontal alignment of the title, use the `hjust` argument. It 
can take values between `0` and `1`. If the value is closer to `0`, the text
will be left-aligned and viceversa.

```{r axis14, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + ggtitle('Diaplacement vs Mileage') +
  theme(plot.title = element_text(hjust = 0.3))
```

## Putting it all together...

### Title

```{r axis15, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + ggtitle('Diaplacement vs Mileage') +
  theme(plot.title = element_text(color = 'blue', family = 'Arial',
    face = 'bold', size = 12, hjust = 0.5))
```

### X axis label

```{r axis16, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + xlab('Diaplacement') +
  theme(axis.title.x = element_text(color = 'blue', family = 'Arial',
    face = 'bold', size = 8, hjust = 0.5, angle = 15))
```

### Y axis label

```{r axis17, fig.align='center', message=FALSE, warning=FALSE}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) + ylab('Miles Per Gallon') +
  theme(axis.title.y = element_text(color = 'blue', family = 'Arial',
    face = 'italic', size = 8, vjust = 0.3, angle = 90))
```