# Text Annotations {#textann}

## Introduction

In this chapter, we will learn to add text to the plots.

- add custom text 
- modify color
- modify size
- modify fontface
- modify angle

## Annotate

We will use the `annotate()` function to add custom text to the plots. You can 
use the `annotate()` function to add rectangles/segments/pointrange as well but 
our focus will be on adding text. Let us start with a simple scatter plot.

```{r ann1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

`annotate()` takes the following arguments:

- `geom`     : specify text
- `x`        : x axis location
- `y`        : y axis location
- `label`    : custom text 
- `color`    : color of text
- `size`     : size of text
- `fontface` : fontface of text
- `angle`    : angle of text

### Add Text

Let us begin by adding text to a scatter plot. We will use the mtcars data set
and continue to examine the relationship between displacement and miles per 
gallon. To add the text, we have to indicate that we are using `annotate()` 
for adding text, and we do this by ensuring that the first input is the word 
`'text'`. Now, ggplot2 knows that it should add a text to the plot but it still
needs other information such as:

- where should the text appear on the plot i.e. location of the text
- and the text itself

We will provide the location by specifying points on the X and Y axis which are
also the second and third inputs to `annotate()` and the final input is the text
itself, which in our example is `'Sample Text'`.

```{r ann2, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  annotate('text', x = 200, y = 30, label = 'Sample Text')
```

### Color

Now that we know how to add text, let us look at modifying the appearance of 
the text. To change the color, use the `color` argument. In the below example,
we modify the color to `'red'`.

```{r ann4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  annotate('text', x = 200, y = 30, label = 'Sample Text', color = 'red')
```

### Size

The `size` argument can be used to modify the size of the text.

```{r ann5, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  annotate('text', x = 200, y = 30, label = 'Sample Text', size = 6)
```

### Font

To choose a font of your liking, use the `font` argument and supply it a valid
value.

```{r ann6, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  annotate('text', x = 200, y = 30, label = 'Sample Text', fontface = 'bold')
```

### Angle

The angle of the text can also be modified using the `angle` argument. In the 
below example, we modify the angle of the text to `25`.

```{r ann7, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  annotate('text', x = 200, y = 30, label = 'Sample Text', angle = 25)
```

## Putting it all together..

```{r ann3, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  annotate('text', x = 200, y = 30, label = 'Sample Text',
           color = 'red', size = 6, fontface = 'bold', angle = 25)
```
