# Scatter Plots {#scatter}

## Introduction

In this chapter, we will: 

- build scatter plots
- modify point
    - color
    - fill
    - alpha
    - shape
    - size
- fit regression line

```{r scat1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

## Basic Plot

As we did in the previous chapter, let us begin by creating a scatter plot using  
`geom_point()` to examine the relationship between displacement and miles per 
gallon using the mtcars data. 

```{r scat2, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg))
```

## Jitter

If you want to avoid over plotting, use the `position` argument and supply it 
the value `'jitter'`. It adds random noise to a plot and makes it easier to 
read.

```{r scat22, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg), position = 'jitter')
```

Another way to avoid over plotting is to use `geom_jitter()`.

```{r scat3, fig.align='center'}
ggplot(mtcars) +
  geom_jitter(aes(disp, mpg))
```

## Aesthetics

Now let us modify the appearance of the points. There are two ways:

- specify values
- map them to variables using `aes()`

### Specify Values 

#### Color

To modify the color of the points, you can use the  `color` argument and 
supply it a valid color name. In the below example, we change the color of the
points to `'blue'`. Keep in mind that the `color` argument should be outside
`aes()`.

```{r scat6, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg), color = 'blue', position = 'jitter')
```

#### Alpha

The transparency of the color can be modified using the `alpha` argument. It
takes values between `0` and `1`.

```{r scat7, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg), color = 'blue', alpha = 0.4, position = 'jitter')
```

#### Shape

The shape of the points can be modified using the `shape` argument. It
takes values between `0` and `25`.

```{r scat9, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg), shape = 3, position = 'jitter')
```

#### Size

The size of the points can be modified using the `size` argument. It can take 
any value greater than `0`.

```{r scat11, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg), size = 3, position = 'jitter')
```

### Map Variables 

So far, we have specified values for color, shape, size etc. Now, let us map 
them to variables using `aes()`.

#### Color

You can modify the color of the points by mapping them to a variable using 
`aes()`. It allows you to examine the relationship between two continuous
variables at different levels of a categorical variable. 

```{r scat4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl)), 
             position = 'jitter')
```

The color can be mapped to a conitnuous variable as well and in this case you
will be able to examine the relationship betweem two continuous variable for 
a range of value of a third variable. 

```{r scat5, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp), 
             position = 'jitter')
```

#### Shape

Shape can be mapped to categorical variables. In the below example, we use 
`factor()` to convert `cyl` to categorical data before mapping shape to it.
ggplot2 will throw an error if you map shape to a continuous variable. 

```{r scat8, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl)), position = 'jitter')
```

#### Size

Size must be always mapped to continuous variables. In the below example, we
have mapped size to `hp` variable. 

```{r scat10, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp), color = 'blue', position = 'jitter')
```

If you map size to categorical data as shown in the below example, ggplot2 
will throw a warning.

```{r scat10a, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = factor(cyl)), color = 'blue', position = 'jitter')
```

## Regression Line

`geom_smooth()` allows us to fit a regression line to the plot. By default it 
will use least squares method to fit the line but you can also use the loess 
method. In the below example, we fit a regression line using the least squares 
technique by supplying the value `'lm'` to the `method` argument.

```{r aes12a, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point(position = 'jitter') +
  geom_smooth(method = 'lm', se = FALSE)
```

The `se` argument will add a confidence interval around the regression line, 
if set to `TRUE`.

#### Conf. Interval

```{r aes13a, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point(position = 'jitter') +
  geom_smooth(method = 'lm', se = TRUE)
```

#### Loess Method

In the below example, we use the loess method instead of the default least 
squares method to fit the regression line.

```{r aes14a, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point(position = 'jitter') +
  geom_smooth(method = 'loess', se = FALSE)
```

### Intercept & Slope

If you know the intercept and the slope of the line, you can use `geom_abline()`. 
Let us regress `mpg` over `disp` and then use the result to add the line.

```{r aes17a}
lm(mpg ~ disp, data = mtcars)
```

### Add Line

```{r aes17c, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point(position = 'jitter') +
  geom_abline(slope = -0.04122, intercept = 29.59985)
```

