# Line Graphs {#ggplot2-line-graph}

## Introduction

In this chapter, we will learn to:

- build 
    - simple line chart
    - grouped line chart
- map aesthetics to variables
- modify line
    - color
    - type
    - size

```{r line, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Case Study

We will use a data set related to GDP growth rate. You can download it from 
[here](https://github.com/rsquaredacademy/datasets/blob/master/gdp.csv). It 
contains GDP (Gross Domestic Product) growth data for the BRICS (Brazil, 
Russia, India, China, South Africa) for the years 2000 to 2005.

### Data

```{r import200, eval=TRUE, message=FALSE}
gdp <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/gdp.csv')
gdp
```

## Line Chart

To create a line chart, use `geom_line()`. In the below example, we examine the
GDP growth rate trend of India for the years 2000 to 2005.

```{r line100, fig.align='center'}
ggplot(gdp, aes(year, india)) +
  geom_line()
```

### Line Color

To modify the color of the line, use the `color` argument and supply it a valid
color name. In the below example, we modify the color of the line to `'blue'`. 
Remember that the `color` argument should be outside `aes()`.

```{r line1, fig.align='center'}
ggplot(gdp, aes(year, india)) +
  geom_line(color = 'blue')
```

### Line Type

The line type can be modified using the `linetype` argument. It can take 7 different
values. You can specify the line type either using numbers or words as shown below:

- 0 : blank
- 1 : solid
- 2 : dashed
- 3 : dotted
- 4 : dotdash
- 5 : longdash
- 6 : twodash

Let us modify the line type to dashed style by supplying the value `2` to the
`linetype` argument.

```{r line2, fig.align='center'}
ggplot(gdp, aes(year, india)) +
  geom_line(linetype = 2)
```

The above example can be recreated by supplying the value `'dashed'` instead 
of `2`.

```{r line3, fig.align='center'}
ggplot(gdp, aes(year, india)) +
  geom_line(linetype = 'dashed')
```

### Line Size

The width of the line can be modified using the `size` argument. It can take 
any value above 0 as input.

```{r line4, fig.align='center'}
ggplot(gdp, aes(year, india)) +
  geom_line(size = 2)
```

## Multiple Lines

### Modify Data

Now let us map the aesthetics to the variables. The data used in the above 
example cannot be used as we need a variable with country names. We will use 
`gather()` function from the `tidyr` package to reshape the data.

```{r line5, fig.align='center'}
gdp2 <- gdp %>% 
  select(year, growth, india, china) %>% 
  gather(key = country, value = gdp, -year)

gdp2
```

In the original data, to plot GDP trend of multiple countries we will have to 
use `geom_line()` multiple times. But in the reshaped data, we have the 
country names as one of the variables and this can be used along with the 
`group` argument to plot data of multiple countries with a single line of code 
as shown below. By mapping country to the `group` argument, we have plotted 
data of all countries.

```{r line6, fig.align='center'}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line()
```

In the above plot, we cannot distinguish between the lines and there is no way
to identify which line represents which country. To make it easier to identify
the trend of different countries, let us map the `color` argument to the 
variable country as shown below. Now, each country will be represented by line 
of different color.

```{r line7, fig.align='center'}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line(aes(color = country))
```

We can map `linetype` argument to country as well. In this case, each country
will be represented by a different line type.

```{r line8, fig.align='center'}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line(aes(linetype = country))
```

We can map the width of the line to the variable country as well. But in this 
case, the plot does not look either elegant or intuitive. 

```{r line9, fig.align='center'}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line(aes(size = country))
```

Remember that in all the above cases, we mapped the arguments to a variable 
inside `aes()`.
