# Bar Plots {#ggplot2-bar-plot}

## Introduction

In this chapter, we will learn to:

- build 
    - simple bar plot
    - stacked bar plot
    - grouped bar plot
    - proportional bar plot
- map aesthetics to variables
- specify values for 
    - bar color
    - bar line color
    - bar line type
    - bar line size

```{r bar, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

## Data

```{r import201, eval=TRUE, message=FALSE}
ecom <- read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/ecom.csv',
  col_types = list(col_factor(levels = c('Desktop', 'Mobile', 'Tablet')), 
  col_logical(), col_logical(), 
  col_factor(levels = c('Affiliates', 'Direct', 'Display', 'Organic', 'Paid', 'Referral', 'Social'))))
ecom
```

### Data Dictionary

- device: device used to visit the website
- bouncers: whether visit was a bouncer (exit website from landing page)
- purchase: whether visitor purchased
- referrer: referrer website/search engine


## Basic Plot

We can create a bar plot using `geom_bar()`. It takes a single input, a 
categorical variable. In the below example, we plot the number of visits for 
each device type. 

```{r bar2, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device))
```

## Bar Color

The color of the bars can be modified using the `fill` argument. In the below
example, we assign different colors to the 3 bars in the plot. If you use the
`color` argument, it will modify the color of the bar line and not the 
background color of the bars. We will look at that later in the chapter.

```{r bar3, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device), fill = c('red', 'blue', 'green'))
```

## Stacked Bar Plot

If you want to look at distribution of one categorical variable across the 
levels of another categorical variable, you can create a stacked bar plot. In
ggplot2, a stacked bar plot is created by mapping the `fill` argument to the
second categorical variable. In the below example, we have mapped `fill` to 
`referrer` variable. 

```{r bar7, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device, fill = referrer))
```

## Grouped Bar Plot

Grouped bar plots are a variation of stacked bar plots. Instead of being 
stacked on top of one another, the bars are placed next to one another and 
grouped by levels. In the below example, we create a grouped bar plot and you
can observe that the bars are placed next to one another instead of being 
stacked as was shown in the previous example. To create a grouped bar plot,
use the `position` argument and set it to `'dodge'`. 

```{r bar8, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device, fill = referrer), position = 'dodge')
```

## Proportional Bar Plot

In a proportional bar plot, the height of all the bars is proportional or same. 
To create a proportional bar plot, use the `position` argument and set it to 
`'fill'`.

```{r bar10, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device, fill = referrer), position = 'fill')
```

## Horizontal Bar Plot

A horizontal bar plot can be created by flipping the coordinate axes of a 
regular plot. To flip the axes, use `coord_flip()` as shown below.

```{r bar9, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device, fill = referrer)) +
  coord_flip()
```

## Bar Line 

The color of the bar line can be modified using the `color` argument. The color
can be specified either using its name or hex code.

```{r bar4, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device), fill = 'white', color = c('red', 'blue', 'green'))
```

To modify the line type of the bar line, use the `linetype` argument. It can 
take values between `0` and `6`.

```{r bar5, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device), fill = 'white',  color = 'black', linetype = 2)
```

The width of the bar line can be modified using the `size` argument. It can
take any value greater than `0`.

```{r bar6, fig.align='center'}
ggplot(ecom) +
  geom_bar(aes(device), fill = 'white', color = 'black', size = 2)
```