# Modify Axis {#ggplot2-modify-axis}

In this chapter, we will learn how to modify the X and Y axis using the 
following functions:

- Continuous Axis
    - `scale_x_continuous()`
    - `scale_y_continuous()`

- Discrete Axis
    - `scale_x_discrete()`
    - `scale_y_discrete()`


## Continuous Axis

If the X and Y axis represent continuous data, we can use 
`scale_x_continuous()` and `scale_y_continuous()` to modify the axis. They take
the following arguments:

- name
- limits
- breaks
- labels
- position

Let us continue with the scatter plot we have used in previous chapter.

```{r guide2, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg))
```

The `name` argument is used to modify the X axis label. In the below example, 
we change the X axis label to `'Displacement'`. In previous chapters, we 
have used `xlab()` to work with the X axis label.

```{r guide3, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  scale_x_continuous(name = "Displacement")
```

To modify the range, use the `limits` argument. It takes a vector of length
2 i.e. 2 values, the lower and upper limit of the range. It is an alternative
for `xlim()`.

```{r guide4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  scale_x_continuous(limits = c(0, 600))
```

In the above plot, the ticks on the X axis appear at `0`, `200`, `400` and 
`600`. Let us say we want the ticks to appear more closer i.e. the difference 
between the tick should be reduced by `50`. The `breaks` argument will allow
us to specify where the ticks appear. It takes a numeric vector equal to the 
length of the number of ticks.

```{r guide5, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  scale_x_continuous(breaks = c(150, 300, 450))
```

We can change the tick labels using the `labels` argument. In the below 
example, we use words instead of numbers. When adding labels, we need to
ensure that the length of the `breaks` and `labels` are same.

```{r guide6, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  scale_x_continuous(breaks = c(150, 300, 450),
    labels = c('One Hundred Fifty', 'Three Hundred', 'Four Hundred Fifity'))
```

The position of the axes can be changed using the `position` argument. In the 
below example, we can move the axes to the top of the plot by supplying the 
value `'top'`.

```{r guide7, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  scale_x_continuous(position = 'top')
```

### Putting it all together..

```{r guide8, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg)) +
  scale_x_continuous(name = "Displacement", limits = c(0, 600),
                     breaks = c(0, 150, 300, 450, 600), position = 'top',
                     labels = c('0', '150', '300', '450', '600'))
```

### Y Axis - Continuous

```{r guide9, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg)) +
  scale_y_continuous(name = "Miles Per Gallon", limits = c(0, 45),
                     breaks = c(0, 15, 30, 45), position = 'right',
                     labels = c('0', '15', '30', '45'))
```

## Discrete Axis

If the X and Y axis represent discrete or categorical data, `scale_x_discrete()` 
and `scale_y_discrete()` can be used to modify them. They take the following arguments:

- name
- labels
- breaks
- position

The above options serve the same purpose as in the case of continuous scales.

### Axis Label

```{r guide11, fig.align='center'}
ggplot(mtcars) +
  geom_bar(aes(factor(cyl))) +
  scale_x_discrete(name = "Number of Cylinders")
```

### Axis Tick Labels

```{r guide12, fig.align='center'}
ggplot(mtcars) +
  geom_bar(aes(factor(cyl))) +
  scale_x_discrete(labels = c("4" = "Four", "6" = "Six", "8" = "Eight"))
```

### Axis Breaks

```{r guide13, fig.align='center'}
ggplot(mtcars) +
  geom_bar(aes(factor(cyl))) +
  scale_x_discrete(breaks = c("4", "6", "8"))
```

### Axis Position

```{r guide14, fig.align='center'}
ggplot(mtcars) +
  geom_bar(aes(factor(cyl))) +
  scale_x_discrete(position = 'bottom')
```

## Putting it all together...

```{r guide15, fig.align='center'}
ggplot(mtcars) + geom_bar(aes(factor(cyl))) +
  scale_x_discrete(name = "Number of Cylinders",
    labels = c("4" = "Four", "6" = "Six", "8" = "Eight"),
    breaks = c("4", "6", "8"), position = "bottom")
```