# Faceting {#ggplot2-facets}

## Introduction

In this chapter, we will learn about faceting i.e. combining plots.

```{r facet, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

Let us continue with the scatter plot examining the relationship between 
displacement and miles per gallon but let us make one additional change.
We now want 3 sub plots for each type of cylinder. How can we do this? We can
split or group the data by cylinder type and plot the subset of data which 
means dealing with 3 different data sets, plotting 3 plots and arranging them
for comparison. ggplot2 offers the following 2 functions which allow us to plot
subset of data with a simple formula based interface:

- `facet_grid()`
- `facet_wrap()`

Faceting allows us to create multiple sub plots. It partitions a plot into a 
matrix of panels with each panel showing a different subset of data.

## Grid

### Vertical

`facet_grid()` allows us to split up the data by one or two discrete variables 
and create sub plots. The sub plots can be arranged horizontally or vertically 
using a formula of the form `vertical ~ horizontal`. In the below example, 3 
sub plots are created, one each for the levels of the `cyl` variable and 
the sub plots are arranged vertically


```{r facet2, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) + 
  geom_point() +
  facet_grid(cyl ~ .)
```

### Horizontal

Below we reproduce the previous example but arrange the sub plots horizontally.

```{r facet1, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) + 
  geom_point() +
  facet_grid(. ~ cyl)
```

### Vertical & Horizontal

In certain cases, we might want different discrete variables to represent the
horizontal and vertical direction. In the below example, we examine the 
relationship between displacement and miles per gallon for different combinations
of `cyl` and `gear` variables. 

```{r facet3, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) + 
  geom_point() +
  facet_grid(cyl ~ gear)
```

Below, we switch the variables representing the vertical and horizontal 
directions.

```{r facet4, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) + 
  geom_point() +
  facet_grid(gear ~ cyl)
```

### Scales

If you carefully observe the second example, the range of `X` axis is same for
all the 3 sub plots i.e. it is a fixed range. You can allow each of the sub 
plots to have different range using the `scales` argument and supplying it the
value `'free'`. 

```{r facet5, fig.align='center'}
ggplot(mtcars, aes(disp, mpg, color = factor(cyl))) +
  geom_point() + 
  facet_grid(. ~ cyl, scales = "free")
```

Now, each of the sub plot has a different range.

### Switch Labels

In the third example, the labels are displayed at the bottom for `X` axis and 
at the right for the `Y` axis. It can be changed using the `switch` argument 
and supplying the value `'both'`. The labels will now be displayed at the top 
for the `X` axis and at left for the `Y` axis. If you just want to change the 
labels for a particular axis, use the values `x` and `y` for the `X` and `Y` 
axis respectively.

```{r facet6, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) + 
  geom_point() + 
  facet_grid(cyl ~ gear, switch = "both") 
```

## Wrap

`facet_wrap()` allows us to arrange sub plots in a certain number of rows and 
columns. In the below example, we will use `facet_wrap()` to arrange the sub 
plots in a single row.

```{r facet7, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  facet_wrap(~cyl)
```

### Specify Rows

To arrange the sub plots in a specific number of rows, use the `nrow` argument.
In the below example, we arrange the sub plots in 2 rows.

```{r facet8, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  facet_wrap(~cyl, nrow = 2)
```

### Specify Columns

Here, we arrange the sub plots in 3 columns instead of rows using the `ncol`
argument.

```{r facet9, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  facet_wrap(~cyl, ncol = 3)
```

### Scales

You can allow each of the sub plots to have different range using the `scales` 
argument and supplying it the value `'free'`. 

```{r facet12, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  facet_wrap(~cyl, scales = "free")
```

### Rows & Columns

If 2 discrete variables are used to create the sub plots, we can either use
the formula interface to specify the variables as shown below 

```{r facet10, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  facet_wrap(~cyl + gear, nrow = 2)
```

or use a character vector of variable names.

```{r facet11, fig.align='center'}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  facet_wrap(c("cyl", "gear"), ncol = 2)
```