# Modify Legend {#ggplot2-modify-legend}

In this chapter, we will focus on modifying the appearance of legend of plots when 
the aesthetics are mapped to variables. 

## Color

We will learn to modify the following when `color` is mapped to categorical variables:

- title
- breaks
- limits
- labels
- values

```{r leg1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

### Basic Plot {-}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the color of the points
to the `cyl` variable. 

```{r leg2, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl)))
```

As you can see, the legend acts as a guide for the `color` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Values {-}

To change the default colors in the legend, use the `values` argument and
supply a character vector of color names. The number of colors specified 
must be equal to the number of levels in the categorical variable mapped.
In the below example, `cyl` has 3 levels (4, 6, 8) and hence we have specified
3 colors.

```{r leg3, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"))
```

### Title {-}

In the previous example, the title of the legend (`factor(cyl)`) is not very
intuitive. If the user does not know the underlying data, they will not be able
to make any sense out of it. Let us change it to `Cylinders` using the `name`
argument.

```{r leg4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(name = "Cylinders", 
    values = c("red", "blue", "green"))
```

Now, the user will know that the different colors represent number of cylinders in
the car.

### Limits {-}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars with at least 6 cylinders. One way to approach this
would be to filter the data using `filter` from dplyr and then visualize it. 
Instead, we will use the `limits` argument and filter the data for visualization.

```{r leg5, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"), limits = c(6, 8))
```

As you can see above, `ggplot2` returns a warning message indicating data related
to 4 cylinders has been dropped. If you observe the legend, it now represents 
only 4 and 6 cylinders.

### Labels {-}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to `Four`, `Six` and `Eight` in the next example. Ensure that
the labels are intuitive and easy to interpret for the end user of the plot.

```{r leg6, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'))
```

### Breaks {-}

When there are large number of levels in the mapped variable, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
two levels (4, 8) of the mapped variable.

```{r leg7, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    breaks = c(4, 8))
```

### Putting it all together... {-}

```{r leg8, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(name = "Cylinders", values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'), limits = c(4, 6, 8), breaks = c(4, 6, 8))
```

## Fill

we will learn to modify the following using `scale_fill_manual()` when `fill` is mapped to categorical variables:

- title
- breaks
- limits
- labels
- values

### Plot {-}

Let us start with a scatter plot examining the relationship between 
displacement and miles per gallon from the mtcars data set. We will map `fill`
to the `cyl` variable. 

```{r leg9}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22)
```

As you can see, the legend acts as a guide for the `color` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Title {-}

The title of the legend (`factor(cyl)`) is not very intuitive. If the user 
does not know the underlying data, they will not be able to make any sense out 
of it. Let us change it to `Cylinders` using the `name` argument.

```{r leg10}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(name = "Cylinders", 
    values = c("red", "blue", "green"))
```

### Values {-}

To change the default colors in the legend, use the `values` argument and
supply a character vector of color names. The number of colors specified 
must be equal to the number of levels in the categorical variable mapped.
In the below example, `cyl` has 3 levels (4, 6, 8) and hence we have specified
3 colors.

```{r leg11}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"))
```

### Labels {-}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to `Four`, `Six` and `Eight` in the next example. Ensure that
the labels are intuitive and easy to interpret for the end user of the plot.

```{r leg12}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'))
```

### Limits {-}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars with at least 6 cylinders. One way to approach this
would be to filter the data using `filter` from dplyr and then visualize it. 
Instead, we will use the `limits` argument and filter the data for visualization.

```{r leg13}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    limits = c(6, 8))
```

As you can see above, `ggplot2` returns a warning message indicating data related
to 4 cylinders has been dropped. If you observe the legend, it now represents 
only 4 and 6 cylinders.

### Breaks {-}

When there are large number of levels in the mapped variable, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
two levels (4, 8) of the mapped variable.

```{r leg14}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(values = c("red", "blue", "green"),
    breaks = c(4, 8))
```

### Putting it all together... {-}

```{r leg15}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, fill = factor(cyl)), shape = 22) +
  scale_fill_manual(name = "Cylinders", values = c("red", "blue", "green"),
    labels = c('Four', 'Six', 'Eight'), limits = c(4, 6, 8), breaks = c(4, 6, 8))
```

## Shape

We will learn to modify the following using `scale_shape_manual` when `shape` is mapped to categorical variables:

- title
- breaks
- limits
- labels
- values

### Plot {-}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the shape of the points
to the `cyl` variable. 

```{r leg16, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl)))
```

As you can see, the legend acts as a guide for the `shape` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Title {-}

The title of the legend (`factor(cyl)`) is not very intuitive. If the user does 
not know the underlying data, they will not be able to make any sense out of it. 
Let us change it to `Cylinders` using the `name` argument.

```{r leg17, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(name = "Cylinders", values = c(4, 12, 24))
```

If you have mapped shape/size to a discrete variable which has less than six 
categories, you can use `scale_shape()`.

```{r leg18, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape(name = 'Cylinders')
```

### Values {-}

To change the default shapes in the legend, use the `values` argument and
supply a numeric vector of shapes. The number of shapes specified 
must be equal to the number of levels in the categorical variable mapped.
In the below example, `cyl` has 3 levels (4, 6, 8) and hence we have specified
3 different shapes.

```{r leg19, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 12, 24))
```

### Labels {-}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to `Four`, `Six` and `Eight` in the next example. Ensure that
the labels are intuitive and easy to interpret for the end user of the plot.

```{r leg20, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 12, 24), labels = c('Four', 'Six', 'Eight'))
```

### Limits {-}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars with at least 6 cylinders. One way to approach this
would be to filter the data using `filter` from dplyr and then visualize it. 
Instead, we will use the `limits` argument and filter the data for visualization.


```{r leg21, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 24), limits = c(6, 8))
```

As you can see above, `ggplot2` returns a warning message indicating data related
to 4 cylinders has been dropped. If you observe the legend, it now represents 
only 4 and 6 cylinders.

### Breaks {-}

When there are large number of levels in the mapped variable, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
two levels (4, 8) of the mapped variable.

```{r leg22, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 12, 24), breaks = c(4, 8))
```

### Putting it all together... {-}

```{r leg23, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(name = "Cylinders", labels = c('Six', 'Eight'),  
     values = c(4, 24), limits = c(6, 8), breaks = c(6, 8))
```

## Size

We will learn to modify the following using `scale_size_continuous` when `size` aesthetic is mapped to variables:

- title
- breaks
- limits
- range
- labels
- values

### Plot {-}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the size of the points
to the `hp` variable. Remember, `size` must always be mapped to a continuous 
variable.

```{r leg24}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp))
```

As you can see, the legend acts as a guide for the `size` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Title {-}

The title of the legend (`hp`) is not very intuitive. If the user does 
not know the underlying data, they will not be able to make any sense out of it. 
Let us change it to `Horsepower` using the `name` argument.

```{r leg25}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(name = "Horsepower")
```

### Range {-}

The range of the size of points can be modified using the `range` argument. We
need to specify a lower and upper range using a numeric vector. In the below
example, we use `range` and supply the lower and upper limits as `3` and `6`. 
The size of the points will now lie between `3` and `6` only.

```{r leg26}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(range = c(3, 6))
```

### Limits {-}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars whose horsepower is between `100` and `350`. 
One way to approach this would be to filter the data using `filter` from dplyr 
and then visualize it. Instead, we will use the `limits` argument and filter 
the data for visualization.

```{r leg27}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(limits = c(100, 350))
```

### Breaks {-}

When the range of the variable mapped to size is large, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
certain midpoints (125, 200, 275) of the mapped variable.

```{r leg28}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(breaks = c(125, 200, 275))
```

### Labels {-}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to "1 Hundred", "2 Hundred" and "3 Hundred" in the next example. 
Ensure that the labels are intuitive and easy to interpret for the end user of 
the plot.

```{r leg29}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(breaks = c(100, 200, 300),
    labels = c("1 Hundred", "2 Hundred", "3 Hundred"))
```

### Putting it all together... {-}

```{r leg30}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(name = "Horsepower", range = c(3, 6), 
    limits = c(0, 400), breaks = c(100, 200, 300),
    labels = c("1 Hundred", "2 Hundred", "3 Hundred"))
```

## Transparency 

We will learn to modify the following using `scale_alpha_continuous()` when `alpha` or transparency is mapped to variables:

- title
- breaks
- limits
- range
- labels
- values

### Plot {-}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the transparency of 
the points to the `hp` variable. Remember, `alpha` must always be mapped to a 
continuous variable.

```{r leg31, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue')
```

As you can see, the legend acts as a guide for the `alpha` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Title {-}

The title of the legend (`hp`) is not very intuitive. If the user does 
not know the underlying data, they will not be able to make any sense out of it. 
Let us change it to `Horsepower` using the `name` argument.

```{r leg32, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous("Horsepower")
```

### Breaks {-}

When the range of the variable mapped to size is large, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
certain midpoints (125, 200, 275) of the mapped variable.

```{r leg33, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(breaks = c(125, 200, 275))
```

### Limits {-}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars whose horsepower is between `100` and `350`. 
One way to approach this would be to filter the data using `filter` from dplyr 
and then visualize it. Instead, we will use the `limits` argument and filter 
the data for visualization.

```{r leg34, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(limits = c(100, 350))
```

### Range {-}

The range of the transparency of points can be modified using the `range` 
argument. We need to specify a lower and upper range using a numeric vector. 
In the below example, we use `range` and supply the lower and upper limits as 
`0.4` and `0.8`. The transparency of the points will now lie between `0.4` and 
`0.8` only.

```{r leg35, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(range = c(0.4, 0.8))
```

### Labels {-}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to "1 Hundred", "2 Hundred" and "3 Hundred" in the next example. 
Ensure that the labels are intuitive and easy to interpret for the end user of 
the plot.

```{r leg36, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(breaks = c(100, 200, 300),
    labels = c("1 Hundred", "2 Hundred", 
      "3 Hundred"))
```

### Putting it all together... {-}

```{r leg37, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous("Horsepower", breaks = c(100, 200, 300),
    limits = c(100, 350), range = c(0.4, 0.8),
    labels = c("1 Hundred", "2 Hundred", "3 Hundred"))
```

## Guide

In this section, we will learn to modify 

- title
- label
- and bar

So far, we have learnt to modify the components of a legend using `scale_*` 
family of functions. Now, we will use the `guide` argument and supply it 
values using the `guide_legend()` function.

### Title {-}

#### Title Alignment {-}

The horizontal alignment of the title can be managed using the `title.hjust` 
argument. It can take any value between `0` and `1`.

- 0 (left)
- 1 (right)

In the below example, we align the title to the center by assigning the value
`0.5`.

```{r leg38, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5))
```

#### Title Alignment (Vertical) {-}

To manage the vertical alignment of the title, use `title.vjust`.

```{r leg39, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    title = "Horsepower", title.position = "top", title.vjust = 1))
```

#### Title Position {-}

The position of the title can be managed using `title.posiiton` argument. It 
can be positioned at:

- top
- bottom
- left
- right


```{r leg40, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top"))
```

### Label {-}

#### Label Position {-}

The position of the label can be managed using the `label.position` argument.
It can be positioned at:

- top
- bottom
- left
- right

In the below example, we position the label at right.

```{r leg41, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.position = "right"))
```

#### Label Alignment {-}

The horizontal alignment of the label can be managed using the `label.hjust` 
argument. It can take any value between `0` and `1`.

- 0 (left)
- 1 (right)

In the below example, we align the label to the center by assigning the value
`0.5`.

- alignment 
    - 0 (left)
    - 1 (right)

```{r leg42, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.hjust = 0.5))
```

#### Labels Alignment (Vertical) {-}

The vertical alignment of the label can be managed using the `label.vjust` 
argument.

```{r leg43, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    label.vjust = 0.8))
```

#### Direction {-}

The direction of the label can be either horizontal or veritcal and it can be 
set using the `direction` argument. 

```{r leg44, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(direction = "horizontal"))
```

#### Rows {-}

The label can be spread across multiple rows using the `nrow` argument. In the
below example, the label is spread across 2 rows.

```{r leg45, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(nrow = 2))
```

#### Reverse {-}

The order of the labels can be reversed using the `reverse` argument. We need
to supply logical values i.e. either `TRUE` or `FALSE`. If `TRUE`, the order
will be reversed.

```{r leg46, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(reverse = TRUE))
```

#### Putting it all together... {-}

```{r leg47, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top", label.position = "right", 
      direction = "horizontal", label.hjust = 0.5, nrow = 2, reverse = TRUE)
  )
```

### Legend Bar  {-}

So far we have looked at modifying components of the legend when it acts as a
guide for `color`, `fill` or `shape` i.e. when the aesthetics have been mapped 
to a categorical variable. In this section, you will learn about
`guide_colorbar()` which will allow us to modify the legend when the aesthetics
are mapped to a continuous variable.

#### Plot {-}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the color of the points
to the `hp` variable. 

```{r leg48}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp))
```

#### Width {-}

The width of the bar can be modified using the `barwidth` argument. It is used
inside the `guide_colorbar()` function which itself is supplied to the `guide`
argument of `scale_color_continuous()`. 

```{r leg49, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barwidth = 10))
```

#### Height {-}

Similarly, the height of the bar can be modified using the `barheight` argument.

```{r leg50, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barheight = 3))
```

#### Bins {-}

The `nbin` argument allows us to specify the number of bins in the bar.

```{r leg51, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    nbin = 4))
```

#### Ticks {-}

The ticks of the bar can be removed using the `ticks` argument and setting it 
to `FALSE`.

```{r leg52, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    ticks = FALSE))
```

#### Upper/Lower Limits {-}

The upper and lower limits of the bars can be drawn or undrawn using the 
`draw.ulim` and `draw.llim` arguments. They both accept logical values.

```{r leg53, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    draw.ulim = TRUE, draw.llim = FALSE))
```

#### Guides: Color, Shape & Size

The `guides()` function can be used to create multiple legends to act as a
guide for `color`, `shape`, `size` etc. as shown below. First, we map color, 
shape and size to different variables. Next, in the `guides()` function, we
supply values to each of the above aesthetics to indicate the type of legend.

```{r leg54, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, 
    size = qsec, shape = factor(gear))) + 
  guides(color = "colorbar", shape = "legend", size = "legend")
```

#### Guides: Title {-}

To modify the components of the different legends, we must use the 
`guide_*` family of functions. In the below example, we use `guide_colorbar()`
for the legend acting as guide for color mapped to a continuous variable and
`guide_legend()` for the legends acting as guide for shape/size mapped to 
categorical variables.

```{r leg55, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, size = wt, shape = factor(gear))) + 
  guides(color = guide_colorbar(title = "Horsepower"),
    shape = guide_legend(title = "Weight"), size = guide_legend(title = "Gear")
  )
```