# Themes {#ggplot2-themes}

## Introduction

In this final chapter, we will learn to modify the appearance of all non data 
components of the plot such as:

- axis
- legend
- panel
- plot area
- background
- margin
- facets

```{r load, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(readr)
```

## Basic Plot

We will continue with the scatter plot examining the relationship between 
displacement and miles per gallon from the the mtcars data set.

```{r theme1, fig.align='center'}
p <- ggplot(mtcars) +
  geom_point(aes(disp, mpg))
p
```

## Axis 

### Text

The `axis.title.x` argument can be used to modify the appearance of the `X` 
axis. In the below example, we modify the color and size of the title using
the `element_text()` function. Remember, whenever you are trying to modify the
appearance of a theme element which is a text, you must use `element_text()`.

You can use `axis.title.y` to modify the `Y` axis title and to modify the 
title of both the axis together, use `axis.title`.

```{r theme2, fig.align='center'}
p + theme(axis.title.x = element_text(color = "red", size = 10, face = "italic"))
```

### Ticks

To modify the appearance of the axis ticks, use the `axis.ticks` argument. You can
change the color, size, linetype and length of the ticks using the `element_line()`
function as shown below. 

```{r theme5, fig.align='center'}
p + theme(axis.ticks = element_line(color = 'blue', size = 1.25, linetype = 2), 
          axis.ticks.length = unit(1, "cm"))
```

### Line

The `axis.line` argument should be used to modify the appearance of the 
axis lines. You can change the color, size and linetype of the line using
the `element_line()` function.

```{r theme6, fig.align='center'}
p + theme(axis.line = element_line(color = 'red', size = 1.5, linetype = 3))
```

## Legend

Now, let us look at modifying the non-data components of a legend.

```{r theme7, fig.align='center'}
p <- ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl), shape = factor(gear)))
p
```

### Background 

The background of the legend can be modified using the `legend.background`
argument. You can change the background color, the border color and line type
using `element_rect()`.

```{r theme30, fig.align='center'}
p + theme(legend.background = element_rect(fill = 'gray', linetype = 3,  
          color = "black"))
```

### Text 

The appearance of the text can be modified using the `legend.text` argument.
You can change the color, size and font using the `element_text()` function.

```{r theme34, fig.align='center'}
p + theme(legend.text = element_text(color = 'green', face = 'italic'))
```

### Title 

The appearance fo the title of the legend can be modified using the 
`legend.title` argument. You can change the color, size, font and alignment
using `element_text()`.

```{r theme35, fig.align='center'}
p + theme(legend.title = element_text(color = 'blue', face = 'bold'),
          legend.title.align = 0.1)
```

### Position 

The position and direction of the legend can be changed using:

- `legend.position`
- and `legend.direction`

```{r theme36, fig.align='center'}
p + theme(legend.position = "top", legend.direction = "horizontal")
```

## Themes

### Classic Dark on Light

```{r theme17, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme_bw()
```

### Default Gray

```{r theme18, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme_gray()
```

### Light

```{r theme19, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme_light()
```

### Minimal

```{r theme20, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme_minimal()
```

### Dark

```{r theme21, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme_dark()
```

### Classic

```{r theme22, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme_classic()
```

### Void (Empty)

```{r theme23, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg)) +
  theme_void()
```
