# Export & Communicate {#ggplot2-export-communicate}

In this chapter, we will save figures at print resolution with
`ggsave()` and run a storytelling checklist before sharing.

:::{.callout-note}
## Why this matters

A great on-screen plot that exports blurry or unlabeled fails at the
finish line. Export settings and a final message check are what your
audience actually sees.
:::

:::{.callout-warning}
## Common pitfall

Screenshots of the RStudio plot pane are low resolution and carry UI
chrome. Always export with `ggsave()` for anything you share.
:::

```{r exp-setup, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
```

## Saving with ggsave

`ggsave()` writes the last plot. Set width, height and dpi
explicitly for print-crisp output:

```{r exp1, eval=FALSE}
p <- ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  labs(title = "Displacement vs mileage",
    x = "Displacement (cu. in.)", y = "Miles per gallon",
    color = "Cylinders")

ggplot2::ggsave("displacement-vs-mileage.png", plot = p,
  width = 8, height = 6, dpi = 300)
```

Match dimensions to the destination: 8x6 inches suits slides, while
narrower widths (e.g. 6x4) fit single-column manuscripts.

## Vector Formats

For manuscripts and posters, export PDF or SVG so lines stay sharp at
any zoom:

```{r exp2, eval=FALSE}
ggplot2::ggsave("displacement-vs-mileage.pdf", plot = p, width = 8, height = 6)
ggplot2::ggsave("displacement-vs-mileage.svg", plot = p, width = 8, height = 6)
```

## Storytelling Checklist

Before sharing any figure, run this Knaflic-inspired check:

- **Context:** does the title state the conclusion, not just the topic?
- **Clutter:** are gridlines, borders and decimals reduced to the minimum?
- **Focus:** is the key series emphasized with color while the rest recede?
- **Labels:** are axes, units and sources stated so the figure stands alone?
- **Honesty:** does the axis start at zero for bars, and are missing data disclosed?

## Before and After

Apply the checklist to a default scatter plot: add a message-first
title, direct axis units and a restrained palette.

```{r exp3, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl)), size = 2.5) +
  labs(title = "Heavier engines get fewer miles per gallon",
    x = "Displacement (cu. in.)", y = "Miles per gallon", color = "Cylinders") +
  theme_minimal()
```

## Where to go next {.unnumbered}

- Storytelling with Data (Knaflic) for the full narrative framework
- Claus Wilke, Fundamentals of Data Visualization, for design principles
