# Modify Legend {#ggplot2-modify-legend}

In this chapter, we customize legend guides: titles, labels and colorbars using the guide argument with guide_legend() and guide_colorbar().

:::{.callout-note}
## Why this matters

Guides control title placement, label layout and colorbar sizing for publication-ready legends.
:::

:::{.callout-warning}
## Common pitfall

Over-customizing direction, rows and justification in one guide hurts readability. Change one element at a time.
:::

```{r setup-legend-g, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Customize legends with guides {#customize-legends-with-guides}

In this section, we will learn to modify 

- title
- label
- and bar

So far, we have learnt to modify the components of a legend using `scale_*` 
family of functions. Now, we will use the `guide` argument and supply it 
values using the `guide_legend()` function.

### Change the guide title {#change-the-guide-title .unnumbered}

#### Title Alignment {.unnumbered}

The horizontal alignment of the title can be managed using the `title.hjust` 
argument. It can take any value between `0` and `1`.

- 0 (left)
- 1 (right)

In the below example, we align the title to the center by assigning the value
`0.5`.

```{r legg38, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5))
```

#### Title Alignment (Vertical) {.unnumbered}

To manage the vertical alignment of the title, use `title.vjust`.

```{r legg39, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    title = "Horsepower", title.position = "top", title.vjust = 1))
```

#### Title Position {.unnumbered}

The position of the title can be managed using `title.position` argument. It 
can be positioned at:

- top
- bottom
- left
- right


```{r legg40, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top"))
```

### Change guide labels {#change-guide-labels .unnumbered}

#### Label Position {.unnumbered}

The position of the label can be managed using the `label.position` argument.
It can be positioned at:

- top
- bottom
- left
- right

In the below example, we position the label at right.

```{r legg41, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.position = "right"))
```

#### Label Alignment {.unnumbered}

The horizontal alignment of the label can be managed using the `label.hjust` 
argument. It can take any value between `0` and `1`.

- 0 (left)
- 1 (right)

In the below example, we align the label to the center by assigning the value
`0.5`.

- alignment 
    - 0 (left)
    - 1 (right)

```{r legg42, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(label.hjust = 0.5))
```

#### Labels Alignment (Vertical) {.unnumbered}

The vertical alignment of the label can be managed using the `label.vjust` 
argument.

```{r legg43, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    label.vjust = 0.8))
```

#### Direction {.unnumbered}

The direction of the label can be either horizontal or vertical and it can be 
set using the `direction` argument. 

```{r legg44, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(direction = "horizontal"))
```

#### Rows {.unnumbered}

The label can be spread across multiple rows using the `nrow` argument. In the
below example, the label is spread across 2 rows.

```{r legg45, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(nrow = 2))
```

#### Reverse {.unnumbered}

The order of the labels can be reversed using the `reverse` argument. We need
to supply logical values i.e. either `TRUE` or `FALSE`. If `TRUE`, the order
will be reversed.

```{r legg46, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(reverse = TRUE))
```

#### Putting it all together {.unnumbered}

```{r legg47, fig.align='center'}
ggplot(mtcars) + geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_manual(values = c("red", "blue", "green"),
    guide = guide_legend(title = "Cylinders", title.hjust = 0.5, 
      title.position = "top", label.position = "right", 
      direction = "horizontal", label.hjust = 0.5, nrow = 2, reverse = TRUE)
  )
```

### Change the legend color bar {#change-the-legend-color-bar .unnumbered}

So far we have looked at modifying components of the legend when it acts as a
guide for `color`, `fill` or `shape` i.e. when the aesthetics have been mapped 
to a categorical variable. In this section, you will learn about
`guide_colorbar()` which will allow us to modify the legend when the aesthetics
are mapped to a continuous variable.

#### Plot {.unnumbered}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the color of the points
to the `hp` variable. 

```{r legg48}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp))
```

#### Width {.unnumbered}

The width of the bar can be modified using the `barwidth` argument. It is used
inside the `guide_colorbar()` function which itself is supplied to the `guide`
argument of `scale_color_continuous()`. 

```{r legg49, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barwidth = 10))
```

#### Height {.unnumbered}

Similarly, the height of the bar can be modified using the `barheight` argument.

```{r legg50, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    barheight = 3))
```

#### Bins {.unnumbered}

The `nbin` argument allows us to specify the number of bins in the bar.

```{r legg51, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    nbin = 4))
```

#### Ticks {.unnumbered}

The ticks of the bar can be removed using the `ticks` argument and setting it 
to `FALSE`.

```{r legg52, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    ticks = FALSE))
```

#### Upper/Lower Limits {.unnumbered}

The upper and lower limits of the bars can be drawn or undrawn using the 
`draw.ulim` and `draw.llim` arguments. They both accept logical values.

```{r legg53, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_continuous(guide = guide_colorbar(
    draw.ulim = TRUE, draw.llim = FALSE))
```

#### Guides: Color, Shape & Size

The `guides()` function can be used to create multiple legends to act as a
guide for `color`, `shape`, `size` etc. as shown below. First, we map color, 
shape and size to different variables. Next, in the `guides()` function, we
supply values to each of the above aesthetics to indicate the type of legend.

```{r legg54, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, 
    size = qsec, shape = factor(gear))) + 
  guides(color = "colorbar", shape = "legend", size = "legend")
```

#### Guides: Title {.unnumbered}

To modify the components of the different legends, we must use the 
`guide_*` family of functions. In the below example, we use `guide_colorbar()`
for the legend acting as guide for color mapped to a continuous variable and
`guide_legend()` for the legends acting as guide for shape/size mapped to 
categorical variables.

```{r legg55, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp, size = wt, shape = factor(gear))) + 
  guides(color = guide_colorbar(title = "Horsepower"),
    shape = guide_legend(title = "Gear"), size = guide_legend(title = "Weight")
  )
```

## Exercises

1. Center a legend title with `guide_legend(title.hjust = 0.5)`.
2. Move discrete legend labels with `label.position`.
3. Size a colorbar with `barwidth`/`barheight` for a continuous guide.

Worked solutions in `solutions/legend.md` (attempt first).

## Where to go next {.unnumbered}

- ggplot2 book scales and guides chapters
