# Scales, Formatting & Color Accessibility {#ggplot2-scales-accessibility}

In this chapter, we will format axis labels with the `scales` package,
use colorblind-safe viridis palettes and check contrast for accessibility.

:::{.callout-note}
## Why this matters

Raw numbers like `150000` slow readers down. Formatted labels
(`$150k`, `45%`) and palettes readable by colorblind viewers make
charts instantly legible to everyone.
:::

:::{.callout-warning}
## Common pitfall

Red-green contrasts are invisible to the most common form of
colorblindness. Never encode critical groups with a red-versus-green
scale alone.
:::

```{r sca-setup, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(scales)
```

## Percent Labels

`label_percent()` turns 0–1 fractions into readable percentages:

```{r sca1, fig.align='center'}
df <- data.frame(group = c("A", "B", "C"), share = c(0.45, 0.30, 0.25))
ggplot(df) +
  geom_bar(aes(group, share), stat = "identity", fill = "steelblue") +
  scale_y_continuous(labels = label_percent())
```

## Dollar Labels

`label_dollar()` formats currency axes:

```{r sca2, fig.align='center'}
sales <- data.frame(month = 1:6, revenue = c(12000, 15000, 13500, 18000, 21000, 19500))
ggplot(sales) +
  geom_line(aes(month, revenue), linewidth = 1) +
  geom_point(aes(month, revenue), size = 2) +
  scale_y_continuous(labels = label_dollar())
```

## Comma Labels

`label_comma()` keeps large counts readable:

```{r sca3, fig.align='center'}
counts <- data.frame(year = 2019:2024, users = c(120000, 150000, 148000, 190000, 230000, 265000))
ggplot(counts) +
  geom_bar(aes(factor(year), users), stat = "identity", fill = "steelblue") +
  scale_y_continuous(labels = label_comma())
```

## Viridis Palettes

Viridis palettes are perceptually uniform and readable by most
colorblind viewers. Use them for continuous fills:

```{r sca4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = hp)) +
  scale_color_viridis_c()
```

And for discrete groups:

```{r sca5, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  scale_color_viridis_d()
```

## Contrast Check

Before publishing, verify your figure in grayscale: print it or
desaturate it. If groups are still distinguishable by luminance, the
contrast passes. The `scale_*_viridis_*()` palettes pass this test by
design; hand-picked pastels often fail it.

## Where to go next {.unnumbered}

- scales package reference: formatting functions gallery
- Claus Wilke, Fundamentals of Data Visualization: color chapters
- Colorblind simulator (e.g. Coblis) to check exported figures
