# Aesthetics {#ggplot2-aesthetics}

## Introduction

:::{.callout-note}
## Why this chart

Aesthetics map variables to position, color, size and shape so patterns pop out.
:::

:::{.callout-warning}
## Common pitfall

Putting a constant inside aes() (e.g. aes(color='red')) creates a bogus legend; constants belong outside aes().
:::

In this chapter, we will focus on the aesthetics i.e. color, shape, size, alpha, 
line type, line width etc. We can map these to variables or specify values for 
them. If we want to map the above to variables, we have to specify them within 
the `aes()` function. We will look at both methods in the following sections.

Explore aesthetics such as 

- color
- shape
- size
- fill
- alpha
- width

```{r aes1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
```

## Libraries, Code & Data

We will use the following libraries in this chapter:

- [readr](https://readr.tidyverse.org/)
- [ggplot2](https://ggplot2.tidyverse.org/)
- [dplyr](https://dplyr.tidyverse.org/)
- [tidyr](https://tidyr.tidyverse.org/)

All the data sets used in this chapter can be found [here](https://github.com/rsquaredacademy/datasets) 
and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/a596a4604b2ebda163313caa272f05cd).

### Data

```{r import, eval=TRUE, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
ecom
```

### Data Dictionary

- id: row id
- referrer: referrer website/search engine
- device: device used to visit the website
- bouncers: whether the visit bounced (single page viewed)
- n_visit: number of visits
- n_pages: number of pages visited
- duration: time spent on the website (in seconds)
- country: country of origin
- purchase: whether visitor purchased
- order_items: number of items ordered
- order_value: order value of visitor (in dollars)

## Color

In ggplot2, when we mention `color` or `colour`, it usually refers to the color 
of the geoms. The `fill` argument is used to specify the color of the shapes in 
certain cases. In this first section, we will see how we can specify the color 
for the different geoms we learnt in the previous chapter. 

### Point

For points, the `color` argument specifies the color of the point for certain 
shapes and border for others. The `fill` argument is used to specify the 
background for some shapes and will not work with other shapes. Let us look at 
an example:

```{r aes2, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg, color = factor(cyl))) +
  geom_point()
```

We can map the variable to color in the `geom_point()` function as well since 
it inherits the data from the `ggplot()` function.

```{r aes3, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(aes(color = factor(cyl)))
```

If you do not want to map a variable to color, you can specify it separately 
using the `color` argument but in this case it should be outside the `aes()` 
function.

```{r aes4, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(color = 'blue')
```

Now we will change the shape of the points to understand the difference between 
`color` and `fill` arguments. It can be again mapped to variables or values. 
Let us map shape to variables.

```{r aes5, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg, shape = factor(cyl))) +
  geom_point()
```

Let us map shape to `cyl` in the `geom_point()` function. Remember, when you
are mapping an aesthetic to a variable, it must be inside `aes()`.

```{r aes6, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(aes(shape = factor(cyl)))
```

Instead of mapping shape to a variable, let us specify a value for shape. In 
this case, `shape` is not wrapped inside `aes()` as we are not mapping it to 
a variable.

```{r aes7, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(shape = 5)
```

Let us specify a color for the point using `color` argument.

```{r aes8, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(shape = 5, color = 'blue')
```

Background color cannot be added for all shapes. In the below example, we try 
to modify the background color using the `fill` argument but it does not work.

```{r aes9, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(shape = 5, fill = 'blue')
```

Since the shape number is now greater than 21, `fill` argument will add background color 
in the below case.

```{r aes10, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(shape = 22, fill = 'blue')
```

In shapes greater than number 21, `color` argument will modify the border of the shape.

```{r aes11, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(shape = 22, color = 'blue')
```

Let us map size of points to a variable. It is advised to map size only to 
continuous variables and not categorical variables.

```{r aes12, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg, size = disp)) +
  geom_point()
```

If you map size to categorical variables, ggplot2 will throw a warning.

```{r aes13, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(size = 4)
```

To modify the opacity of the color, use the `alpha` argument.

```{r aes14, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(aes(alpha = factor(cyl)), color = 'blue')
```

## Line Chart

So far we have focussed on `geom_point()` to learn how to map aesthetics to 
variables. To explore line type and line width, we will use `geom_line()`. In 
the previous chapter, we used `geom_line()` to build line charts. Now we will 
modify the appearance of the line. In the section below, we will specify values 
for color, line type and width. In the next section, we will map the same to 
variables in the data. We will use a new data set. You can download it from 
[here](https://github.com/rsquaredacademy/datasets/blob/master/gdp.csv). It 
contains GDP (Gross Domestic Product) growth data for the BRICS (Brazil, 
Russia, India, China, South Africa) for the years 2000 to 2005.

### Data

```{r import2a, eval=TRUE, message=FALSE}
gdp <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/gdp.csv')
```

A line chart can be created using `geom_line()`. In the below example, we 
examine the GDP trend of India and modify the color of the line to `'blue'`.

```{r aes15, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp, aes(year, india)) +
  geom_line(color = 'blue')
```

To modify the line type, use the `linetype` argument. It can take values 
between 1 and 5. 

```{r aes16, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp, aes(year, india)) +
  geom_line(linetype = 2)
```

The line type can also be mentioned in the following way:

```{r aes17, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp, aes(year, india)) +
  geom_line(linetype = 'dashed')
```

The width of the line can be modified using the `linewidth` argument.

```{r aes18, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp, aes(year, india)) +
  geom_line(linewidth = 2)
```

Now let us map the aesthetics to the variables. The data used in the above 
example cannot be used as we need a variable with country names. We will use 
`pivot_longer()` function from the `tidyr` package to reshape the data.

```{r aes19, fig.align='center', fig.height=5, fig.width=6}
gdp2 <-
  gdp |>
  select(year, growth, india, china) |>
  tidyr::pivot_longer(cols = -year, names_to = "country", values_to = "gdp")

gdp2
```

To map the aesthetics to a variable, we must use the `group` argument. In the 
below example, we map the aesthetics to `country`. But we cannot distinguish 
between the lines as their color, width and line type are the same. We have 
easily plotted the GDP trend of all countries using the `group` argument. Now, 
let us ensure that we can distinguish and identidy them using different 
aesthetics.

```{r aes20, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line()
```

Let us begin by ensuring that the lines have different color using the
`color` argument within `aes()` and assigning it the variable `country`.

```{r aes21, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line(aes(color = country))
```

Instead of color, now we modify the line type using the `linetype` argument.

```{r aes22, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line(aes(linetype = country))
```

In the below instance, we assign different width to the lines using the `linewidth`
argument.

```{r aes23, fig.align='center', fig.height=5, fig.width=6}
ggplot(gdp2, aes(year, gdp, group = country)) +
  geom_line(aes(linewidth = country))
```

Before we wrap up, let us quickly see how we can map aesthetics to variables 
for different plots.

## Bar Plots

Here we create a stacked bar plot by mapping `fill` to `purchase`.

```{r aes24, fig.align='center', fig.height=5, fig.width=6}
ggplot(ecom, aes(device, fill = purchase)) +
  geom_bar()
```

## Histograms

Instead of a bar chart, we create a histogram and again map `fill` to 
`purchase`.

```{r aes25, fig.align='center', fig.height=5, fig.width=6}
ggplot(ecom) +
  geom_histogram(aes(duration, fill = purchase), bins = 10)
```

## Box Plots

We repeat the same exercise below, but replace the bar plot with a box plot.

```{r aes26, fig.align='center', fig.height=5, fig.width=6}
ggplot(ecom) +
  geom_boxplot(aes(device, duration, fill = purchase))
```

In all the above cases, you can observe that when we are mapping aesthetics 
such as color, fill, shape, size or linetype to variables, they are all wrapped 
inside `aes()`.

## Exercises

1. Map `color` to cylinders and set a constant point `size = 3` outside `aes()`.
2. Map `linewidth` (not `size`) to a line geom and explain the difference.
3. Reshape a two-column wide table with `pivot_longer()` and map the key to color.

Worked solutions in `solutions/aesthetics.md` (attempt first).

## Where to go next {.unnumbered}

- ggplot2 book (ggplot2-book.org) for theory behind this layer
- Claus Wilke, Fundamentals of Data Visualization, for perception guidance
