# Finale: Publication Polish with patchwork, ggrepel & ggtext {#ggplot2-finale-polish}

In this chapter, we will combine multi-plot layouts (`patchwork`),
collision-free labels (`ggrepel`) and rich text (`ggtext`) into three
polished before/after case studies.

:::{.callout-note}
## Why this matters

Individual plots answer single questions. Combining them into one
composed figure with clean labels is what separates an exploratory
chart from a published one.
:::

:::{.callout-warning}
## Common pitfall

Each extension adds load time and new syntax. Master core ggplot2
first; reach for these only for final figures.
:::

```{r fin-setup, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(patchwork)
library(ggrepel)
library(ggtext)
```

## Case 1: Multi-Plot Layout

Before: two related plots shown separately. After: one composed
figure with a shared title.

```{r fin1, fig.align='center'}
p1 <- ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  labs(title = "Displacement vs mileage")
p2 <- ggplot(mtcars) +
  geom_boxplot(aes(factor(cyl), mpg, fill = factor(cyl))) +
  labs(title = "Mileage by cylinders")
p1 + p2 + plot_annotation(title = "Engine size explains fuel economy")
```

Stack vertically with `/` and collect legends with
`plot_layout(guides = "collect")`:

```{r fin2, fig.align='center'}
p1 / p2 + plot_layout(guides = "collect")
```

## Case 2: Collision-Free Labels

Before: `geom_text()` labels pile onto each other. After:
`geom_text_repel()` pushes them apart with thin leader lines.

```{r fin3, fig.align='center'}
cars <- mtcars
cars$model <- rownames(mtcars)
efficient <- cars[cars$mpg > 30, ]
ggplot(cars) +
  geom_point(aes(disp, mpg)) +
  ggrepel::geom_text_repel(data = efficient, aes(disp, mpg, label = model), size = 3)
```

## Case 3: Rich Text Titles

Before: plain titles. After: `element_markdown()` renders color and
emphasis inline.

```{r fin4, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, color = factor(cyl))) +
  labs(title = "Heavier engines drink <b style='color:#d62728'>more fuel</b>") +
  theme(plot.title = ggtext::element_markdown())
```

## Where to go next {.unnumbered}

- patchwork reference for nested and tagged layouts
- ggrepel and ggtext references for label-heavy figures
- Export chapter to save the composed figure at print resolution
