# Position & Stats Tuning {#ggplot2-position-stats}

In this chapter, we will fine-tune how ggplot2 arranges and transforms
marks: parametric position adjustments, summary statistics and modern
stat mapping syntax.

:::{.callout-note}
## Why this matters

Defaults get you 80% there. Tuning `position_dodge()` widths,
`position_jitter()` spreads and `stat_summary()` aggregations takes a
plot from roughly right to exactly right.
:::

:::{.callout-warning}
## Common pitfall

String shorthands like `position = "dodge"` hide their parameters.
Use the `position_*()` function form whenever you need to tune width,
height or stacking behavior.
:::

```{r pos-setup, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
```

## Position Dodge

Grouped bar plots dodge side by side. Control the gap with `width` —
smaller values widen the gap between groups.

```{r pos1, fig.align='center'}
ggplot(mtcars) +
  geom_bar(aes(factor(cyl), fill = factor(gear)), position = position_dodge(width = 0.8))
```

## Position Jitter

Jitter spreads overlapping points. Set `width` to spread horizontally
and `height = 0` when the y axis is exact and must not be distorted.

```{r pos2, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(factor(cyl), mpg), position = position_jitter(width = 0.2, height = 0))
```

`geom_jitter()` is a shortcut with the same arguments:

```{r pos3, fig.align='center'}
ggplot(mtcars) +
  geom_jitter(aes(factor(cyl), mpg), width = 0.2, height = 0)
```

## Position Stack

`position_stack()` piles bars (the default for `geom_bar()`).
Use it explicitly when combining stacked bars with labels:

```{r pos4, fig.align='center'}
ggplot(mtcars) +
  geom_bar(aes(factor(cyl), fill = factor(gear)), position = position_stack())
```

## Stat Summary

`stat_summary()` aggregates y values per x group on the fly. Here we
plot mean horsepower per cylinder count:

```{r pos5, fig.align='center'}
ggplot(mtcars) +
  stat_summary(aes(factor(cyl), hp), fun = mean, geom = "bar", fill = "steelblue")
```

Add spread with `fun.min` and `fun.max`:

```{r pos6, fig.align='center'}
ggplot(mtcars) +
  stat_summary(aes(factor(cyl), hp), fun = mean, fun.min = min, fun.max = max,
    geom = "pointrange", color = "steelblue")
```

## After Stat

Computed statistics must be mapped with `after_stat()`. The old
`..count..` syntax is deprecated. Here density is mapped on a histogram:

```{r pos7, fig.align='center'}
ggplot(mtcars) +
  geom_histogram(aes(mpg, y = after_stat(density)), bins = 10,
    fill = "steelblue", color = "white", linewidth = 0.5)
```

## Where to go next {.unnumbered}

- ggplot2 book (ggplot2-book.org): position and statistical summaries chapters
- Claus Wilke, Fundamentals of Data Visualization: visualizing distributions
