# Data Reshaping for ggplot2 {#ggplot2-reshaping}

In this chapter, we will learn the one idea that unlocks faceting,
grouped aesthetics and multi-series line graphs: long versus wide data.

:::{.callout-note}
## Why this matters

ggplot2 maps one column to one aesthetic. Wide tables (one column per
series) must be reshaped to long tables (one key column plus one value
column) before color, fill or facets can separate the series.
:::

:::{.callout-warning}
## Common pitfall

`gather()` appears in older answers but is retired. Always use
`tidyr::pivot_longer()` in new code.
:::

```{r rsh-setup, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(tidyr)
```

## Wide Data

Wide data keeps each series in its own column. Here quarterly revenue
for two regions sits in `north` and `south`:

```{r rsh1}
wide <- data.frame(
  quarter = c("Q1", "Q2", "Q3", "Q4"),
  north = c(120, 150, 135, 180),
  south = c(100, 130, 160, 175)
)
wide
```

Plotting this directly forces one layer per series — repetitive and
unscalable.

## Long Data

`pivot_longer()` collapses the series columns into a `region` key
column and a `revenue` value column:

```{r rsh2}
long <- tidyr::pivot_longer(wide, cols = c(north, south),
  names_to = "region", values_to = "revenue")
long
```

## Plotting Long Data

One mapping now draws both series, distinguished by color:

```{r rsh3, fig.align='center'}
ggplot(long) +
  geom_line(aes(quarter, revenue, color = region, group = region), linewidth = 1) +
  geom_point(aes(quarter, revenue, color = region), size = 2)
```

The same long shape powers facets and dodged bars with no extra
reshaping.

## Modern Pipe

Chain reshaping into plotting with the native pipe `|>`:

```{r rsh4, fig.align='center'}
wide |>
  tidyr::pivot_longer(cols = c(north, south), names_to = "region", values_to = "revenue") |>
  ggplot() +
  geom_bar(aes(region, revenue, fill = quarter), stat = "identity",
    position = position_dodge(width = 0.8))
```

## Where to go next {.unnumbered}

- tidyr `pivot_longer()` reference for multi-column specs
- Line Graphs chapter for multi-series line recipes
- Faceting chapter for small multiples of long data
