# Quick Tour {#ggplot2-quicktour}

## Introduction

:::{.callout-note}
## Why this chart

A 5-minute on-ramp: one ggplot() plus one geom_*() draws your first chart.
:::

:::{.callout-warning}
## Common pitfall

Skipping labs() early produces cryptic default titles; label axes from the start.
:::

In this chapter, we will quickly build a set of plots that are routinely
used to explore data using `ggplot()` combined with a `geom_*()` layer.
Every plot follows the same recipe: map variables to aesthetics in
`aes()` and add the geometry that draws them.

:::{.callout-note}
`qplot()` was deprecated in ggplot2 3.4.0. It still appears in older
tutorials, but all examples in this book use `ggplot() + geom_*()`,
which is the supported syntax going forward.
:::

```{r gg1, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
```

## Libraries, Code & Data

We will use the following libraries in this chapter:

- [readr](https://readr.tidyverse.org/)
- [ggplot2](https://ggplot2.tidyverse.org/)

All the data sets used in this chapter can be found [here](https://github.com/rsquaredacademy/datasets)
and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/3b25fa07c60d44d561819d2c6ab77978).

## Scatter Plot

Scatter plots are used to examine the relationship between two continuous
variables. The relationship can be examined across the levels of a categorical
variable as well. Let us begin by creating scatter plots. Map the X and Y
variables inside `aes()` and add `geom_point()`.

```{r gg2, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point()
```

If you want the relationship between the two variables to be represented by
both points and a line, add both layers.

```{r gg3, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  geom_line()
```

The color of the points can be mapped to a categorical variable, in our case
`cyl`, using the `color` aesthetic. Ensure that the variable is categorical using
`factor()`.

```{r gg4, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(disp, mpg, color = factor(cyl))) +
  geom_point()
```

The shape and size of the points can also be mapped to variables using the
`shape` and `size` aesthetics as shown in the below examples.

```{r gg5, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(disp, mpg, shape = factor(cyl))) +
  geom_point()
```

Ensure that size is mapped to a continuous variable.

```{r gg6, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(disp, mpg, size = qsec)) +
  geom_point()
```

## Bar Plot

A bar plot represents data in rectangular bars. The length of the bars are
proportional to the values they represent. Bar plots can be either horizontal
or vertical. The X axis of the plot represents the levels or the categories and
the Y axis represents the frequency/count of the variable.

To create a bar plot, map a categorical variable to `x`. You can
convert a variable to type `factor` (R equivalent of categorical) using the
`factor()` function, and add `geom_bar()`.

```{r gg7, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(factor(cyl))) +
  geom_bar()
```
You can create a stacked bar plot by mapping another categorical variable to
`fill`.

```{r gg8, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(factor(cyl), fill = factor(am))) +
  geom_bar()
```

## Box Plot

The box plot is a standardized way of displaying the distribution of data
based on the five number summary: minimum, first quartile, median, third
quartile, and maximum. Box plots are useful for detecting outliers and for
comparing distributions. It shows the shape, central tendancy and variability
of the data.

Box plots are created with `geom_boxplot()`. Map a categorical variable to
`x` and a continuous variable to `y`.

```{r gg9, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_boxplot()
```

To plot the distribution of a single continuous variable without any grouping,
map only `y`. Modern ggplot2 accepts a single-variable boxplot natively.

```{r gg10, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(y = mpg)) +
  geom_boxplot()
```

:::{.callout-tip}
Older code (including earlier editions of this book) used
`aes(x = factor(1))` as a dummy grouping variable for single boxplots.
That workaround is no longer needed but still renders, so you may
encounter it in the wild.
:::

## Line Chart

Line charts are used to examing trends across time. Map the time variable to
`x`, the measurement to `y`, and add `geom_line()`.

```{r gg12, fig.align='center', fig.height=5, fig.width=6}
ggplot(economics, aes(date, unemploy)) +
  geom_line()
```

The appearance of the line can be modified by setting the `color` outside
`aes()` as shown below.

```{r gg13, fig.align='center', fig.height=5, fig.width=6}
ggplot(economics, aes(date, unemploy)) +
  geom_line(color = 'red')
```

## Histogram

A histogram is a plot that can be used to examine the shape and spread of
continuous data. It looks very similar to a bar graph and can be used to detect
outliers and skewness in data. Map the continuous variable to `x` and add
`geom_histogram()`, controlling detail with `bins`.

```{r gg14, fig.align='center', fig.height=5, fig.width=6}
ggplot(mtcars, aes(mpg)) +
  geom_histogram(bins = 5)
```

## Exercises

1. Draw a `ggplot(mtcars)` scatter of `disp` vs `mpg` with `geom_point()`.
2. Add color by cylinders and a `labs()` title.
3. Explain in one sentence why `qplot()` is legacy code.

Worked solutions in `solutions/quicktour.md` (attempt first).

## Where to go next {.unnumbered}

- ggplot2 book (ggplot2-book.org) for theory behind this layer
- Claus Wilke, Fundamentals of Data Visualization, for perception guidance
