# Legend: Shape Size and Alpha {#ggplot2-legend-shape-size-alpha}

In this chapter, we modify legends when shape, size or alpha is mapped to variables.

:::{.callout-note}
## Why this matters

Shape, size and alpha encode extra variables without new panels; map size and alpha to continuous variables only.
:::

:::{.callout-warning}
## Common pitfall

Mapping size or alpha to a discrete variable produces a misleading legend.
:::

```{r setup-legend-cf, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Shape

We will learn to modify the following using `scale_shape_manual` when `shape` is mapped to categorical variables:

- title
- breaks
- limits
- labels
- values

### Plot {.unnumbered}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the shape of the points
to the `cyl` variable. 

```{r legs16, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl)))
```

As you can see, the legend acts as a guide for the `shape` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Title {.unnumbered}

The title of the legend (`factor(cyl)`) is not very intuitive. If the user does 
not know the underlying data, they will not be able to make any sense out of it. 
Let us change it to `Cylinders` using the `name` argument.

```{r legs17, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(name = "Cylinders", values = c(4, 12, 24))
```

If you have mapped shape/size to a discrete variable which has less than six 
categories, you can use `scale_shape()`.

```{r legs18, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape(name = 'Cylinders')
```

### Values {.unnumbered}

To change the default shapes in the legend, use the `values` argument and
supply a numeric vector of shapes. The number of shapes specified 
must be equal to the number of levels in the categorical variable mapped.
In the below example, `cyl` has 3 levels (4, 6, 8) and hence we have specified
3 different shapes.

```{r legs19, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 12, 24))
```

### Labels {.unnumbered}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to `Four`, `Six` and `Eight` in the next example. Ensure that
the labels are intuitive and easy to interpret for the end user of the plot.

```{r legs20, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 12, 24), labels = c('Four', 'Six', 'Eight'))
```

### Limits {.unnumbered}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars with at least 6 cylinders. One way to approach this
would be to filter the data using `filter` from dplyr and then visualize it. 
Instead, we will use the `limits` argument and filter the data for visualization.


```{r legs21, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 24), limits = c(6, 8))
```

As you can see above, `ggplot2` returns a warning message indicating data related
to 4 cylinders has been dropped. If you observe the legend, it now represents 
only 6 and 8 cylinders.

### Breaks {.unnumbered}

When there are large number of levels in the mapped variable, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
two levels (4, 8) of the mapped variable.

```{r legs22, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(values = c(4, 12, 24), breaks = c(4, 8))
```

### Putting it all together... {.unnumbered}

```{r legs23, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, shape = factor(cyl))) +
  scale_shape_manual(name = "Cylinders", labels = c('Six', 'Eight'),  
     values = c(4, 24), limits = c(6, 8), breaks = c(6, 8))
```

## Size

We will learn to modify the following using `scale_size_continuous` when `size` aesthetic is mapped to variables:

- title
- breaks
- limits
- range
- labels
- values

### Plot {.unnumbered}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the size of the points
to the `hp` variable. Remember, `size` must always be mapped to a continuous 
variable.

```{r legs24}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp))
```

As you can see, the legend acts as a guide for the `size` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Title {.unnumbered}

The title of the legend (`hp`) is not very intuitive. If the user does 
not know the underlying data, they will not be able to make any sense out of it. 
Let us change it to `Horsepower` using the `name` argument.

```{r legs25}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(name = "Horsepower")
```

### Range {.unnumbered}

The range of the size of points can be modified using the `range` argument. We
need to specify a lower and upper range using a numeric vector. In the below
example, we use `range` and supply the lower and upper limits as `3` and `6`. 
The size of the points will now lie between `3` and `6` only.

```{r legs26}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(range = c(3, 6))
```

### Limits {.unnumbered}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars whose horsepower is between `100` and `350`. 
One way to approach this would be to filter the data using `filter` from dplyr 
and then visualize it. Instead, we will use the `limits` argument and filter 
the data for visualization.

```{r legs27}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(limits = c(100, 350))
```

### Breaks {.unnumbered}

When the range of the variable mapped to size is large, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
certain midpoints (125, 200, 275) of the mapped variable.

```{r legs28}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(breaks = c(125, 200, 275))
```

### Labels {.unnumbered}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to "1 Hundred", "2 Hundred" and "3 Hundred" in the next example. 
Ensure that the labels are intuitive and easy to interpret for the end user of 
the plot.

```{r legs29}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(breaks = c(100, 200, 300),
    labels = c("1 Hundred", "2 Hundred", "3 Hundred"))
```

### Putting it all together... {.unnumbered}

```{r legs30}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, size = hp)) +
  scale_size_continuous(name = "Horsepower", range = c(3, 6), 
    limits = c(0, 400), breaks = c(100, 200, 300),
    labels = c("1 Hundred", "2 Hundred", "3 Hundred"))
```

## Transparency 

We will learn to modify the following using `scale_alpha_continuous()` when `alpha` or transparency is mapped to variables:

- title
- breaks
- limits
- range
- labels
- values

### Plot {.unnumbered}

Let us start with a scatter plot examining the relationship between displacement
and miles per gallon from the mtcars data set. We will map the transparency of 
the points to the `hp` variable. Remember, `alpha` must always be mapped to a 
continuous variable.

```{r legs31, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue')
```

As you can see, the legend acts as a guide for the `alpha` aesthetic. Now, let 
us learn to modify the different aspects of the legend.

### Title {.unnumbered}

The title of the legend (`hp`) is not very intuitive. If the user does 
not know the underlying data, they will not be able to make any sense out of it. 
Let us change it to `Horsepower` using the `name` argument.

```{r legs32, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous("Horsepower")
```

### Breaks {.unnumbered}

When the range of the variable mapped to size is large, you may not 
want the labels in the legend to represent all of them. In such cases, we can 
use the breaks argument and specify the labels to be used. In the below case, 
we use the `breaks` argument to ensure that the labels in legend represent 
certain midpoints (125, 200, 275) of the mapped variable.

```{r legs33, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(breaks = c(125, 200, 275))
```

### Limits {.unnumbered}

Let us assume that we want to modify the data to be displayed i.e. instead of 
examining the relationship between mileage and displacement for all cars, we 
desire to look at only cars whose horsepower is between `100` and `350`. 
One way to approach this would be to filter the data using `filter` from dplyr 
and then visualize it. Instead, we will use the `limits` argument and filter 
the data for visualization.

```{r legs34, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(limits = c(100, 350))
```

### Range {.unnumbered}

The range of the transparency of points can be modified using the `range` 
argument. We need to specify a lower and upper range using a numeric vector. 
In the below example, we use `range` and supply the lower and upper limits as 
`0.4` and `0.8`. The transparency of the points will now lie between `0.4` and 
`0.8` only.

```{r legs35, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(range = c(0.4, 0.8))
```

### Labels {.unnumbered}

The labels in the legend can be modified using the `labels` argument. Let us 
change the labels to "1 Hundred", "2 Hundred" and "3 Hundred" in the next example. 
Ensure that the labels are intuitive and easy to interpret for the end user of 
the plot.

```{r legs36, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous(breaks = c(100, 200, 300),
    labels = c("1 Hundred", "2 Hundred", 
      "3 Hundred"))
```

### Putting it all together... {.unnumbered}

```{r legs37, fig.align='center'}
ggplot(mtcars) +
  geom_point(aes(disp, mpg, alpha = hp), color = 'blue') +
  scale_alpha_continuous("Horsepower", breaks = c(100, 200, 300),
    limits = c(100, 350), range = c(0.4, 0.8),
    labels = c("1 Hundred", "2 Hundred", "3 Hundred"))
```


## Exercises

1. Map `shape` to cylinders with `scale_shape_manual(values = ...)`.
2. Control point spread with `scale_size_continuous(range = ...)`.
3. Explain why `size`/`alpha` must map to continuous variables.

Worked solutions in `solutions/legend-shape-size-alpha.md` (attempt first).

## Where to go next {.unnumbered}

- ggplot2 book scales and guides chapters
