# Label Polishing & Wrapping {#ggplot2-label-polish}

In this chapter, we will polish text that overflows: wrapping long
axis labels and legend entries with `stringr` and `scales`.

:::{.callout-note}
## Why this matters

Long category names overlap and wreck otherwise clean charts. Wrapped
labels keep every category readable without shrinking fonts to
microscopic sizes.
:::

:::{.callout-warning}
## Common pitfall

Angling text 90 degrees feels like a fix but forces readers to tilt
their heads. Wrap first; rotate only when wrapping cannot save the
layout.
:::

```{r lab-setup, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(ggplot2)
library(stringr)
library(scales)
```

## The Overflow Problem

Long labels collide on crowded axes:

```{r lab1, fig.align='center'}
df <- data.frame(
  program = c("Engineering and Technology", "Business Administration",
    "Health and Medical Sciences", "Arts and Humanities"),
  students = c(320, 280, 240, 190)
)
ggplot(df) +
  geom_bar(aes(program, students), stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Wrapping with stringr

`str_wrap()` inserts line breaks at a fixed width:

```{r lab2, fig.align='center'}
ggplot(df) +
  geom_bar(aes(stringr::str_wrap(program, width = 15), students),
    stat = "identity", fill = "steelblue") +
  labs(x = "Program")
```

## Wrapping Inside Scales

`label_wrap()` wraps at the scale level, keeping the underlying data
untouched:

```{r lab3, fig.align='center'}
ggplot(df) +
  geom_bar(aes(program, students), stat = "identity", fill = "steelblue") +
  scale_x_discrete(labels = scales::label_wrap(15))
```

## Wrapping Long Titles

The same helper tidies wrapped plot titles and subtitles:

```{r lab4, fig.align='center'}
ggplot(df) +
  geom_bar(aes(stringr::str_wrap(program, width = 15), students),
    stat = "identity", fill = "steelblue") +
  labs(x = "Program",
    title = stringr::str_wrap("Student enrollment across four academic programs in 2024", width = 40))
```

## Where to go next {.unnumbered}

- stringr reference for wider text manipulation
- Themes chapter for font sizes that pair with wrapped labels
